\documentclass[fleql]{beamer}

\usetheme[
  progressbar=frametitle,
  numbering=fraction,
  sectionpage=none,
  subsectionpage=none
]{metropolis}

\title{Week 5 \\ Understanding under Threat}
\subtitle{Quantum Mechanics and Artificial Intelligence}
\author{Hans Halvorson}
\date{May 2026}

\usepackage[backend=biber, style=authortitle]{biblatex}
\addbibresource{refs.bib}
\setlength{\bibitemsep}{0.5em}
\AtBeginBibliography{\scriptsize}

\usepackage{amsmath,amssymb}
\usepackage{soul}
\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}

\newcommand{\kw}[1]{\textbf{#1}}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\bra}[1]{\langle #1|}
\newcommand{\ip}[2]{\langle #1|#2\rangle}

\setbeamertemplate{section in toc}[sections numbered]
\setbeamerfont{section in toc}{size=\small}

\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents[currentsection,hideallsubsections,part=\thepart]
  \end{frame}
}

\setbeamertemplate{footline}[frame number]
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{part page}{}

\setcounter{secnumdepth}{1}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Today's plan}

  \textbf{Part I: Quantum Mechanics} \hfill \emph{(lectures 1--2)}
  \begin{itemize}\small
  \item The framework of QM: states, observables, dynamics
  \item The measurement problem and the Copenhagen myth
  \item Interpretations: GRW, Bohm, Everett, \ldots
  \item Complementarity: Bohr's response to the crisis
  \item Bell's theorem and the limits of causal understanding
  \end{itemize}

  \vspace{0.8em}
  \textbf{Part II: Artificial Intelligence} \hfill \emph{(lecture 3)}
  \begin{itemize}\small
  \item AI in science: prediction without understanding?
  \item Content to be developed from notes
  \end{itemize}

\end{frame}

%%% ============================================================
%%% PART I: QUANTUM MECHANICS
%%% ============================================================

\part{Quantum Mechanics}

\begin{frame}{Group Discussion}

  From last week, quantum mechanics is our recurring example of
  a theory where prediction and understanding seem to come apart.

  \vspace{0.8em}
  \begin{enumerate}
  \item Do you feel that you \textbf{understand} quantum mechanics?
    What would it even mean to understand it?
  \item Is ``shut up and calculate'' a satisfying attitude toward a
    fundamental theory of nature?
  \end{enumerate}

\end{frame}

\section{Introduction}

\begin{frame}{Two perspectives today}

  \begin{description}
  \item[First-order] Which interpretation of QM is correct? What is
    the theory \emph{actually} saying about the world?
  \item[Second-order] There has been no consensus on the foundations
    of QM for \emph{100 years}. Should we be concerned? What does
    this tell us about science as a knowledge-producing enterprise?
  \end{description}

  \vspace{0.8em}
  Both questions bear directly on scientific understanding --- which
  is why De Regt and Dieks chose quantum non-locality as their
  motivating example.

\end{frame}

\begin{frame}{The foundational crisis: 1900--1927}

  \begin{itemize}
  \item 1900: Planck introduces the quantum of action $h$
  \item 1905: Einstein explains the photoelectric effect using photons
  \item 1913: Bohr's model of the hydrogen atom
  \item 1925--26: Heisenberg's matrix mechanics; Schr\"{o}dinger's
    wave mechanics
  \item 1927: Born's probability interpretation; Heisenberg's
    uncertainty principle; Bohr's complementarity
  \end{itemize}

  \vspace{0.5em}
  \small The early pioneers were acutely aware that they did
  \emph{not} understand what was going on. Bohr thought our
  ability to understand the physical world was genuinely up
  for grabs.

\end{frame}

\begin{frame}{Apparent resolution: the Copenhagen school}

  \begin{itemize}
  \item By 1927 a working consensus formed around Bohr's school
  \item Bohr's key insight: the existence of the \kw{quantum of
    action} calls for a \emph{revision} of what it means to give an
    objective, unambiguous description of physical phenomena
    \begin{itemize}
    \item An unambiguous description must specify the experimental
      arrangement; the conditions of observation become part of
      the description itself
    \item Classical concepts remain indispensable for describing
      the apparatus, but cannot all be simultaneously applied to
      the quantum object
    \end{itemize}
  \item This leads to \kw{complementarity}: a structural feature of
    unambiguous description, not a limitation on knowledge
  \item This is a positive philosophical proposal, not a
    counsel of despair
  \end{itemize}

\end{frame}

\begin{frame}{Revival of the crisis}

  \begin{itemize}
  \item 1952: Bohm publishes a working hidden-variable
    interpretation --- shattering the apparent consensus
  \item 1957: Everett proposes the relative-state (many-worlds)
    interpretation
  \item 1964: Bell proves his non-locality theorem
  \item Today: active foundational debate; the cutting edge of
    physics proceeds without settling the foundations
  \end{itemize}

  \vspace{0.5em}
  \textbf{The second-order question:} is 100 years without consensus
  acceptable? Does it reveal something deep about QM, or is it a
  sociological failure of physics?

\end{frame}

\section{Framework of Quantum Mechanics}

\begin{frame}{States as vectors}

  \begin{itemize}
  \item The state of a quantum system is a unit vector $\ket{\psi}$
    in a complex Hilbert space $\mathcal{H}$
  \item \kw{Superposition principle}: if $\ket{\psi}$ and $\ket{\phi}$
    are states, so is $\alpha\ket{\psi} + \beta\ket{\phi}$
    (after normalization)
  \item For a \kw{qubit} (two-level system): $\mathcal{H} = \mathbb{C}^2$
  \item Computational basis: $\ket{0} \equiv \ket{\uparrow}$,\quad
    $\ket{1} \equiv \ket{\downarrow}$
  \end{itemize}

\end{frame}

\begin{frame}{The Bloch sphere}

\begin{center}
\begin{tikzpicture}[scale=1.1]
  \draw[thick] (0,0) circle (1.5);
  \draw[gray!50] (1.5,0) arc (0:180:1.5 and 0.45);
  \draw[gray!50,dashed] (-1.5,0) arc (180:360:1.5 and 0.45);
  \draw[->,semithick] (0,-1.9) -- (0,2.1) node[above] {\scriptsize$z$};
  \draw[->,semithick] (-1.9,0) -- (2.2,0) node[right] {\scriptsize$x$};
  \draw[->,semithick,dashed] (0,0) -- (-0.85,-0.6)
    node[below left] {\scriptsize$y$};
  \fill (0,1.5) circle (2pt)
    node[right,font=\scriptsize] {$\ket{0} = \ket{\uparrow}_z$};
  \fill (0,-1.5) circle (2pt)
    node[right,font=\scriptsize] {$\ket{1} = \ket{\downarrow}_z$};
  \fill (1.5,0) circle (1.5pt)
    node[above right,font=\scriptsize] {$\ket{+}_x$};
  \fill (-1.5,0) circle (1.5pt)
    node[above left,font=\scriptsize] {$\ket{-}_x$};
  \draw[->,thick,blue!75] (0,0) -- (0.9,1.1)
    node[right,font=\scriptsize] {$\ket{\psi}$};
  \fill[blue!75] (0,0) circle (1.5pt);
\end{tikzpicture}
\end{center}

\vspace{-0.3em}
\small Every pure state of a qubit is a point on the sphere.
Antipodal points are orthogonal.

\end{frame}

\begin{frame}{Observables as operators}

  \begin{itemize}
  \item Physical quantities: \kw{Hermitian operators} $A = A^\dagger$
  \item Possible measurement outcomes: eigenvalues $\lambda_i$
  \item Eigenstates $\ket{a_i}$: states with definite value $\lambda_i$
  \end{itemize}

  \vspace{0.5em}
  \textbf{Spin operators (Pauli matrices $\times\,\hbar/2$):}
  \[
  S_z = \frac{\hbar}{2}\begin{pmatrix}1&0\\0&-1\end{pmatrix},\quad
  S_x = \frac{\hbar}{2}\begin{pmatrix}0&1\\1&0\end{pmatrix},\quad
  S_y = \frac{\hbar}{2}\begin{pmatrix}0&-i\\i&0\end{pmatrix}
  \]

\end{frame}

\begin{frame}{Superpositions and the puzzle}

  If the system is in $\ket{\psi} = \alpha\ket{a_1}+\beta\ket{a_2}$,
  observable $A$ has \emph{no definite value} --- yet measurement
  always yields a definite outcome $\lambda_1$ or $\lambda_2$.

  \vspace{0.5em}
  \textbf{The puzzle:} what is the system doing between preparation
  and measurement? Does the question even make sense?

  \vspace{0.5em}
  This tension is the engine of every interpretation of QM.

\end{frame}

\begin{frame}{Uncertainty relations}

  Non-commuting observables cannot both have definite values:
  \[
  [S_y,\,S_z] = i\hbar S_x
  \quad\Longrightarrow\quad
  \Delta S_y \cdot \Delta S_z \;\geq\;
  \tfrac{\hbar}{2}|\langle S_x\rangle|
  \]

  \textbf{Spin-$z$ eigenstate decomposed in spin-$y$ basis:}
  \[
  \ket{\uparrow}_z \;=\;
  \tfrac{1}{\sqrt{2}}\ket{\uparrow}_y + \tfrac{1}{\sqrt{2}}\ket{\downarrow}_y
  \]
  so $\Delta S_y = \hbar/2$ is maximal when $S_z$ is definite.

  \vspace{0.3em}
  Conceptually: \kw{Fourier analysis} --- narrow in one basis,
  spread in the conjugate basis.

\end{frame}

\begin{frame}{Composite systems and entanglement}

  Composite systems: $\mathcal{H}_{AB} =
  \mathcal{H}_A \otimes \mathcal{H}_B$, with
  $x\otimes(y+z) = (x\otimes y)+(x\otimes z)$.

  \vspace{0.3em}
  Not all states are product states. The \kw{singlet state}
  \[
  \ket{\Psi^-} = \tfrac{1}{\sqrt{2}}
  \bigl(\ket{\uparrow\downarrow} - \ket{\downarrow\uparrow}\bigr)
  \]
  cannot be written $\ket{\psi}_A\otimes\ket{\phi}_B$:
  it is \kw{entangled}.

  \vspace{0.5em}
  \textbf{Discussion (2 min):} what features does a system have if
  it is entangled with another? What does it mean for its ``state''?

\end{frame}

\begin{frame}{Linear dynamics and the Born rule}

  Time evolution:
  $i\hbar\frac{d}{dt}\ket{\psi} = H\ket{\psi}$,\quad
  solution $\ket{\psi(t)} = U(t)\ket{\psi(0)}$ (\kw{unitary}).

  \vspace{0.3em}
  Crucially, $U$ is \kw{linear}:
  $U(\alpha\ket{\psi}+\beta\ket{\phi}) = \alpha U\ket{\psi}+\beta U\ket{\phi}$.
  This is exactly what generates the measurement problem.

  \vspace{0.5em}
  \textbf{Born rule:}
  \[
  \mathrm{Pr}_\psi(A=\lambda) =
  \langle\psi,\,E^A(\lambda)\,\psi\rangle
  \]
  For $\ket{\psi}=\sum_i c_i\ket{a_i}$:\quad
  $\mathrm{Pr}_\psi(A=\lambda_i) = |c_i|^2$

  \vspace{0.4em}
  \textbf{Discussion:} do we \emph{understand} QM if it is
  just a recipe for computing these probabilities?

\end{frame}

\section{The Measurement Problem}

\begin{frame}{Definition and derivation}

  \textbf{Ideal measurement of $A$:}
  $\ket{a_i}\otimes\ket{r} \;\longrightarrow\; \ket{a_i}\otimes\ket{y_i}$
  (each eigenstate paired with a distinct pointer state --- unitary).

  \vspace{0.5em}
  Now suppose $\ket{\psi} = \sum_i c_i\ket{a_i}$.
  By \textbf{linearity}:
  \[
  \Bigl(\sum_i c_i\ket{a_i}\Bigr)\otimes\ket{r}
  \;\longrightarrow\;
  \sum_i c_i\,\bigl(\ket{a_i}\otimes\ket{y_i}\bigr)
  \]

  The meter ends in an \textbf{entangled superposition} --- no
  definite reading. But we \emph{always} observe a definite outcome.
  \textbf{Contradiction.}

\end{frame}

\begin{frame}{The collapse postulate}

  \textbf{Standard ``solution'':}
  $\sum_i c_i\ket{a_i} \longrightarrow \ket{a_k}$
  with probability $|c_k|^2$.

  \vspace{0.5em}
  \textbf{Problems:}
  \begin{itemize}
  \item Violates unitarity and linearity --- the two most fundamental
    features of the theory
  \item Creates a two-dynamics theory:
    Schr\"{o}dinger evolution \emph{and} collapse
  \item When does collapse occur? Where is the ``Heisenberg cut''?
  \item The theory gives no definition of ``measurement''
  \end{itemize}

\end{frame}

\begin{frame}{The Copenhagen myth}

  \begin{itemize}
  \item Von Neumann (1932) first axiomatized collapse in
    \emph{Grundlagen der Quantenmechanik}; by the 1950s it was
    standard textbook orthodoxy
  \item The label \kw{Copenhagen interpretation} was applied
    retroactively --- Bohr never used it
  \item In reality there was no single Copenhagen view: Bohr,
    Heisenberg, Born, Pauli all held different positions
  \item \textbf{HH:} Bohr never claimed the measurement problem is
    \emph{solved} by collapse --- his actual response is
    philosophically much deeper, and has been systematically misread
  \end{itemize}

\end{frame}

\begin{frame}{Decoherence: not a solution}

  \kw{Decoherence}: environmental interaction entangles the system
  rapidly; the \emph{reduced} state approaches a mixture.

  \vspace{0.5em}
  \textbf{Why this does not solve the measurement problem:}
  \begin{enumerate}
  \item The reduced state is not \emph{uniquely} interpretable as
    a mixture over definite pointer readings
  \item The global state (system $+$ environment) is still a
    superposition --- decoherence is a local approximation,
    not a collapse
  \end{enumerate}

  \vspace{0.3em}
  \small Decoherence explains why interference is hard to observe
  macroscopically. It does not explain why we observe \emph{one}
  outcome.

\end{frame}

\section{Interpretations of Quantum Mechanics}

\begin{frame}{Four strategies}

  \begin{enumerate}
  \item \kw{Reject unitary dynamics}: add stochastic collapse
    \hfill\emph{(GRW)}
  \item \kw{Add hidden variables}: supplement QM with additional
    ontology \hfill\emph{(Bohm)}
  \item \kw{Reject uniqueness of outcomes}: all branches are real
    \hfill\emph{(Everett)}
  \item \kw{Revise the ontological picture}: resist the demand for
    a classical reality behind QM
    \hfill\emph{(relational, QBism, pragmatic)}
  \end{enumerate}

\end{frame}

\begin{frame}{GRW, Bohm, and their problems}

  \textbf{GRW} (Ghirardi, Rimini, Weber 1986): add random
  spontaneous localization events. Free parameters tuned so
  macroscopic objects localize (no pointer superpositions) while
  microscopic systems remain unitary. Open: ontology? Relativistic
  extension?

  \vspace{0.5em}
  \textbf{Bohm (1952)}: both wavefunction and particle positions
  are real; the wave guides particles. Particles always have
  definite positions.
  \begin{itemize}\small
  \item Guidance equation requires a preferred frame ---
    inconsistent with special relativity
  \item Violates the spirit of momentum conservation in
    relativistic interactions
  \end{itemize}

\end{frame}

\begin{frame}{Everett and less ontological interpretations}

  \textbf{Everett (1957)}: unitary dynamics is complete; all
  branches of the wavefunction are real.
  \begin{itemize}\small
  \item \emph{Probability:} deterministic theory --- where do the
    Born-rule probabilities come from?
  \item \emph{Preferred basis:} what selects the ``world'' branches?
  \item No-hidden-variables theorems imply the branches cannot
    themselves be classical
  \end{itemize}

  \vspace{0.5em}
  \textbf{Less ontological} (Rovelli, Fuchs--Schack, Healey):
  resist asking ``what is the wavefunction in itself?''

  \vspace{0.3em}
  \small \textbf{HH:} these approaches are philosophically
  sophisticated but may relocate rather than dissolve the problem.

\end{frame}

\section{Complementarity}

\begin{frame}{Bohr's single-slit thought experiment}

\begin{columns}

\column{0.49\textwidth}
\centering
\textbf{A: bolted screen}
\begin{center}
\begin{tikzpicture}[scale=0.63]
  \fill[gray!55] (0,0.35) rectangle (0.45,2.1);
  \fill[gray!55] (0,-2.1) rectangle (0.45,-0.35);
  \draw[ultra thick] (0.22,2.1) -- (0.22,2.45);
  \draw[ultra thick] (0.22,-2.1) -- (0.22,-2.45);
  \draw[->,thick] (-1.8,0) -- (0,0);
  \draw[->,blue!80,thick] (0.45,0) -- (2.3,0.75);
  \draw[->,blue!50] (0.45,0) -- (2.3,0.15);
  \draw[->,blue!25] (0.45,0) -- (2.3,-0.35);
  \fill[gray!25] (2.3,-2.1) rectangle (2.6,2.1);
  \foreach \y in {0.75,0.15,-0.35} {
    \fill[blue!50] (2.6,\y-0.09) rectangle (3.05,\y+0.09);
  }
\end{tikzpicture}
\end{center}
\small position known $\Rightarrow$ wave picture $\Rightarrow$ \textbf{interference}

\column{0.49\textwidth}
\centering
\textbf{B: free screen}
\begin{center}
\begin{tikzpicture}[scale=0.63]
  \fill[gray!55] (0,0.35) rectangle (0.45,2.1);
  \fill[gray!55] (0,-2.1) rectangle (0.45,-0.35);
  \draw[thick,decorate,
    decoration={zigzag,segment length=4pt,amplitude=2pt}]
    (0.22,-2.1) -- (0.22,-2.7);
  \draw[->,thick] (-1.8,0) -- (0,0);
  \draw[->,blue!80,thick] (0.45,0) -- (2.3,0.45);
  \draw[->,red,thick] (0.22,-1.7) -- (0.22,-2.2)
    node[right,font=\tiny] {recoil};
  \fill[gray!25] (2.3,-2.1) rectangle (2.6,2.1);
  \fill[gray!45] (2.6,-0.1) rectangle (3.05,0.55);
\end{tikzpicture}
\end{center}
\small momentum known $\Rightarrow$ particle picture $\Rightarrow$ \textbf{no interference}

\end{columns}

\end{frame}

\begin{frame}{Complementarity and visualizability}

  Two arrangements are \kw{complementary} when they are mutually
  exclusive and each exhaustive within its domain.

  \vspace{0.5em}
  \begin{description}
  \item[Spacetime description] position, trajectory, ``which path''
    --- \kw{visualizable} (anskuelig) picture
  \item[Causal description] momentum, energy, conservation laws
    --- dynamical picture
  \end{description}

  \vspace{0.5em}
  ``The viewpoint of complementarity presents itself as a rational
  generalization of the very ideal of causality.'' (Bohr 1948)

  \vspace{0.3em}
  \small This connects directly to De Regt \& Dieks: Bohr is
  \emph{not} saying we cannot understand quantum phenomena. He
  is proposing a new kind of understanding that does not require
  a single classical picture.

\end{frame}

\begin{frame}{Bohr's deeper point}

  \begin{itemize}
  \item The quantum of action requires a revision of what counts
    as an \kw{objective description}:
    \begin{itemize}
    \item Objectivity does not disappear --- but it now requires
      specifying the experimental conditions as part of any
      unambiguous statement about a quantum phenomenon
    \item We cannot separate ``how Nature is'' from ``the
      experimental arrangement in which it is observed''
    \end{itemize}
  \item This is not about abandoning objectivity but about
    \emph{reconceiving} it for a domain where the interaction
    between object and instrument cannot be neglected or controlled
  \item The misreading of Bohr as ``shut up and calculate'' has
    distorted the foundations debate for 70 years
  \item \textbf{HH:} Bohr is working in a tradition running
    from Kant through H\o ffding: the conditions of unambiguous
    description are themselves a philosophical achievement
  \end{itemize}

\end{frame}

\section{No-Hidden-Variables Theorems}

\begin{frame}{EPR, von Neumann, and Kochen--Specker}

  \textbf{EPR (1935):} QM is incomplete --- quantum correlations
  must be explained by predetermined hidden variables $\lambda$.

  \vspace{0.5em}
  \textbf{Von Neumann (1932):} no hidden-variable theory can
  reproduce QM given additivity $\mathbb{E}_\lambda(A+B) =
  \mathbb{E}_\lambda(A)+\mathbb{E}_\lambda(B)$ for non-commuting
  $A,B$. Bell (1966): this assumption is physically unmotivated.

  \vspace{0.5em}
  \textbf{Kochen--Specker (1967):} no assignment
  $v(A)\in\mathrm{spec}(A)$ to all observables can satisfy
  functional consistency $v(f(A))=f(v(A))$ and the sum rule for
  commuting pairs. \kw{Value definiteness} is impossible.

  \vspace{0.3em}
  \small Kochen--Specker is more compelling than von Neumann: the
  assumptions are operationally motivated.

\end{frame}

\section{Bell's Theorem}

\begin{frame}{The CHSH scenario}

\begin{center}
\begin{tikzpicture}[scale=0.92]
  \draw[fill=gray!20,thick,rounded corners=3pt]
    (-0.45,-0.4) rectangle (0.45,0.4);
  \node[font=\scriptsize] at (0,0) {source};
  \draw[fill=blue!10,thick,rounded corners=3pt]
    (-4.2,-0.45) rectangle (-2.3,0.45);
  \node[font=\scriptsize] at (-3.25,0) {\textbf{Alice}};
  \node[font=\tiny] at (-3.25,-0.65) {setting: $a$ or $a'$};
  \node[font=\tiny] at (-3.25,0.65) {outcome: $\pm 1$};
  \draw[fill=red!10,thick,rounded corners=3pt]
    (2.3,-0.45) rectangle (4.2,0.45);
  \node[font=\scriptsize] at (3.25,0) {\textbf{Bob}};
  \node[font=\tiny] at (3.25,-0.65) {setting: $b$ or $b'$};
  \node[font=\tiny] at (3.25,0.65) {outcome: $\pm 1$};
  \draw[->,thick] (-0.45,0) -- (-2.3,0);
  \draw[->,thick] (0.45,0) -- (2.3,0);
  \draw[<->,dashed,gray!70] (-2.3,-0.9) -- (2.3,-0.9)
    node[midway,below,font=\tiny,gray!70] {no communication};
\end{tikzpicture}
\end{center}

\vspace{0.3em}
\small The entangled pair is measured far apart. Each party
independently chooses a setting and records a $\pm 1$ outcome.

\end{frame}

\begin{frame}{Local hidden variables and the CHSH inequality}

  Suppose outcomes are determined by $\lambda\sim\rho(\lambda)$,
  with \kw{reality} ($A,B$ definite) and \kw{locality}
  (Alice independent of Bob's setting):
  \[
  E(a,b) = \int A(a,\lambda)\,B(b,\lambda)\,\rho(\lambda)\,d\lambda
  \]

  Under local hidden variables:
  \[
  \bigl|E(a,b) - E(a,b') + E(a',b) + E(a',b')\bigr| \;\leq\; 2
  \]

  \small \textit{Proof sketch:} for each $\lambda$, the combination
  $A(a)[B(b)-B(b')] + A(a')[B(b)+B(b')]$ takes values in
  $\{-2,0,+2\}$.

\end{frame}

\begin{frame}{Quantum violation}

  For $\ket{\Psi^-} = \tfrac{1}{\sqrt{2}}(\ket{\uparrow\downarrow}
  -\ket{\downarrow\uparrow})$:\quad
  $E(a,b) = -\cos(\theta_a-\theta_b)$

  \vspace{0.3em}
  Choose $a=0^\circ$, $a'=90^\circ$, $b=45^\circ$, $b'=135^\circ$:

  \begin{align*}
  E(a,b)   &= -\cos(-45^\circ)  = -\tfrac{1}{\sqrt{2}} \\[3pt]
  E(a,b')  &= -\cos(-135^\circ) = +\tfrac{1}{\sqrt{2}} \\[3pt]
  E(a',b)  &= -\cos(45^\circ)   = -\tfrac{1}{\sqrt{2}} \\[3pt]
  E(a',b') &= -\cos(-45^\circ)  = -\tfrac{1}{\sqrt{2}}
  \end{align*}

  \[
  S = -\tfrac{1}{\sqrt{2}} - \tfrac{1}{\sqrt{2}}
    - \tfrac{1}{\sqrt{2}} - \tfrac{1}{\sqrt{2}}
  = -\tfrac{4}{\sqrt{2}} = -2\sqrt{2}
  \qquad\Rightarrow\qquad
  \boxed{|S| = 2\sqrt{2} \approx 2.83 > 2}
  \]

\end{frame}

\begin{frame}{The Jarrett analysis and what Bell means}

  Jarrett (1984): local hidden variables $=$ two conditions:
  \begin{description}
  \item[PI (parameter independence)] Alice's outcome does not
    depend on Bob's \emph{setting}
  \item[OI (outcome independence)] Alice's outcome does not
    depend on Bob's \emph{outcome}
  \end{description}

  Bell violation $\Leftrightarrow$ at least one of PI or OI fails.
  Bohm/Everett: OI fails. Retrocausal models: PI fails.

  \vspace{0.5em}
  \textbf{HH's focus:} Bell's theorem shows a fundamental
  limitation on \kw{local causal explanation} of quantum
  correlations. Does this mean we \emph{cannot understand} them?

  \vspace{0.3em}
  \small De Regt \& Dieks: understanding does not require causal
  explanation. Bohr: we need a new conceptual framework,
  not a new causal story.

\end{frame}

\section{Conclusion: The QM Crisis}

\begin{frame}{Taking stock}

  \begin{itemize}
  \item QM works to extraordinary precision (electron $g$-factor:
    12 significant figures)
  \item Yet the theory is arguably \emph{inconsistent as stated}
    (the measurement problem), and there is no consensus on how to
    resolve it
  \item Each interpretation pays a significant price: non-locality,
    many worlds, anti-realism, new fundamental dynamics
  \item \textbf{HH:} this is a genuine crisis --- not merely a
    philosophical puzzle we can comfortably set aside
  \end{itemize}

\end{frame}

\begin{frame}{From QM to AI}

  \begin{itemize}
  \item Some physicists say: ``QM works, so who cares if we don't
    understand it?''
  \item If we accept this for QM, we have weakened our ground
    against the parallel argument:
    \begin{quote}\small
    ``AlphaFold predicts protein structures better than any human
    theory --- so who cares if we don't understand it?''
    \end{quote}
  \item \textbf{HH:} I do not think we should be content with this.
    The goal of science is not mere prediction but understanding.
    QM is one thing; the threat from AI is another.
  \end{itemize}

\end{frame}

%%% ============================================================
%%% PART II: AI AND SCIENTIFIC UNDERSTANDING
%%% ============================================================

\part{AI and Scientific Understanding}

\begin{frame}[plain]
  \vfill
  \centering
  {\usebeamerfont{title}\usebeamercolor[fg]{title} Part II}\\[0.5em]
  {\large Artificial Intelligence and Scientific Understanding}
  \vfill
\end{frame}

\section{AI and Scientific Understanding}

\begin{frame}{[Content to be developed]}

  \emph{Notes and content for this section will be added.}

\end{frame}

\nocite{*}

\begin{frame}[allowframebreaks]{Selected References}
\printbibliography[heading=none]
\end{frame}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
