\documentclass[12pt,fleqn]{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\setlength{\parindent}{0em}
\begin{document}

%% Hey, remember next time to PRINT OUT grade sheets

\section*{PHI 201, Final Exam 2025} 

\textbf{Instructions:} Write your name, preceptor's name, and pledge
on the exam booklet. Write your answers \emph{legibly} in the exam
booklet. You have up to three hours to complete the exam, although it
was designed to take no more than two hours. When finished, place your
exam booklet in the box at the front of the room, and leave
quietly.

\section*{A. Engagement}

Please write the following if and only if you were in lecture on
December 1st: ``I attended the lecture on December 1st, and the coupon
code is $x$'', where you replace $x$ with the coupon code.

\section*{B. Translation}

Please translate the following into the symbolism of predicate
logic. Use $Gx$ for ``$x$ is a god'' and $Hxy$ for ``$x$ helps
$y$''. You may assume that quantifiers and variables apply only to
people and gods.

\begin{enumerate}
\item There are at most three gods. 
\item There is a god who helps nobody.   
\item The gods help those who help themselves.
\end{enumerate}

\section*{C. Proofs}

Prove the following sequents. You may simply cite ``prop taut'' for
any substitution instance of a propositional tautology --- no need to
include a proof. You may also substitute quantifier validities and
equivalences, but only if you prove them in your exam booklet.

\begin{enumerate}  
\item $\neg\forall xFx\:\vdash \:\exists x\neg Fx$
\item
  $\forall xFx \leftrightarrow \neg \exists x\exists y Rxy
  \:\vdash\:\exists x\forall y\forall z (Fx \rightarrow \neg Ryz)$
\end{enumerate}  

\section*{D. Models}

Show that the following sequents are invalid by providing countermodels.

\begin{enumerate}
\item $\forall x(Fx \rightarrow Gx),\exists xGx \:\vdash\: \exists xFx$
\item $\forall y\,\exists x\,Rxy,\ \forall x\,\forall y\,\forall z\bigl((Rxy \wedge Rzy)\rightarrow x=z\bigr)
  \;\vdash\; \exists x\,\forall y\,Rxy$
\end{enumerate}

\section*{E. Conceptual}

\emph{Note:} This problem is \emph{extra credit}, and will only be
used to raise grades from A$^-$ to A, or from A to A$^+$.

\medskip Let $\Gamma$ be the set of sentences built from $P$ using
$\neg$ and $\wedge$. Show that for every sentence $\varphi$ in
$\Gamma$, either $P\vdash\varphi$ or $P\vdash\neg\varphi$.



\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
