\documentclass[12pt,fleqn]{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\setlength{\parindent}{0em}
\begin{document}

%% Hey, remember next time to PRINT OUT grade sheets

\section*{PHI 201, Final Exam 2025 (makeup)} 

\textbf{Instructions:} Write your name, preceptor's name, and pledge
on the exam booklet. Write your answers \textbf{legibly}. You have up
to three hours to complete the exam, although it was designed to take
no more than two. You may use one letter-sized page of notes during
the exam.

\section*{A. Engagement}

Please write the following if and only if you were in lecture on
December 1st: ``I attended the lecture on December 1st, and the coupon
code is $x$'', where you replace $x$ with the coupon code. (2 points)

\section*{B. Translation}

Please translate the following into the symbolism of predicate
logic. Use $Gx$ for ``$x$ is a god'' and $Hxy$ for ``$x$ helps
$y$''. You may assume that quantifiers and variables apply only to
people and gods. (4 points each)

\begin{enumerate}
\item There is a god who helps everybody.
\item There are at most three gods. 
\item The gods help those who help themselves.
\end{enumerate}

\section*{C. Proofs}

Prove the following sequents. You may simply cite ``prop taut'' for
any substitution instance of a propositional tautology --- no need to
include a proof. You may also substitute quantifier validities and
equivalences, but only if you prove them in your exam booklet. (6
points each)

\begin{enumerate}  
\item $\neg\forall xFx\:\vdash \:\exists x\neg Fx$
\item
  $\neg\exists x\exists y
  Rxy\leftrightarrow \forall xFx
  \:\vdash\:\exists x\forall y\forall z (Fx \rightarrow \neg Ryz)$
\end{enumerate}

\section*{Please turn over} \newpage 

\section*{D. Models}

Show that the following sequents are invalid by providing
countermodels. (6 points each)

\begin{enumerate}
\item $\forall xFx\to \forall xGx\:\vdash\: \forall x(Fx\to Gx)$
\item
  $\forall y\exists xRxy, \,\forall x\forall y\forall z((Rxy \wedge
  Rzy)\rightarrow x=z ) \:\vdash\: \exists x\forall yRxy$
\end{enumerate}

\section*{E. Conceptual}

\emph{Note:} This problem is \emph{extra credit}, and will only be
used to raise grades from A$^-$ to A, or from A to A$^+$.

\medskip Let $\Gamma$ be the set of sentences built from $P$ using
$\neg$ and $\wedge$. Show that for every sentence $\varphi$ in
$\Gamma$, either $P\vdash\varphi$ or $P\vdash\neg\varphi$.


\section*{The end} 


\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
