\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\title{Logic precept: Week 7}
\date{}
\begin{document}

\maketitle

\vspace*{-4em}

\section*{Review}

\begin{enumerate}
  \item What kinds of sentences are there in predicate logic?
  \item What is the difference between a \textbf{formula} and a \textbf{sentence}?
  \item What is an \textbf{instance} of a universal sentence?
  \item What is the restriction on UI?
\end{enumerate}

\section*{Proofs}

\subsection*{Warmup Problems}

\begin{enumerate}
\item
  $\forall x (F x \rightarrow G x) \ \vdash \ \forall x F x
  \rightarrow \forall x G x$
\item
  $\forall x (P x \rightarrow Q x), \ \forall x (Q x \rightarrow R x)
  \ \vdash \ \forall x (P x \rightarrow R x)$
\item
  $P \rightarrow \forall x F x \ \vdash \ \forall x (P \rightarrow F
  x)$
\item $\forall x \forall y R x y \ \vdash \ \forall y \forall x R x y$
  
\end{enumerate}

\subsection*{Pset Problems}

\begin{enumerate}
\item
  $\forall x (F x \rightarrow \forall y G y) \ \vdash \ \forall x
  \forall y (F x \rightarrow G y)$
\item
  $\forall x \forall y (F x \rightarrow G y) \ \vdash \ \forall x (F x
  \rightarrow \forall y G y)$
\item $\vdash \ \forall x (\forall y R x y \rightarrow R x x)$
\end{enumerate}

\section*{Translation}

\subsection*{Exercise}

How do you symbolize the following?

\begin{enumerate}
  \item All $F$ are $G$.
  \item No $F$ are $G$.
  \item Some $F$ are $G$.
  \item Some $F$ are not $G$.
  \end{enumerate}

\subsection*{Exercise}

Use $F$ for ``is French'', $G$ for ``is German'', $C$ for ``is
Canadian'', $L x y$ for ``$x$ likes $y$'', $a$ for Alice, and $b$ for
Bob. How would you symbolize:

\begin{enumerate}
\item Alice likes Canadians.
\item Alice likes Bob only if Bob likes Canadians.
\item Alice likes Bob only if he likes her.
\item Alice is a German who likes Canadians.
\item Alice is French only if she doesn't like Canadians.
\item Alice likes only those people who don't like Canadians.
\item Someone likes only those people who like Canadians.
\item French people only like Canadians who don't like Germans.
\item Some French people like only those Germans who don't like
  themselves.
\end{enumerate}

\end{document}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
